---
title: "Required packages"
date: November 2023
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will use the lab to **get started**. We will set up R for the course, including installing main libraries and loading the project data set. In **Let's get started (1) and (2)**, a list of the R packages you need for exercises in Manuela's, Zhi's and Maiju's lectures and the project work will be installed.

## Let's get started (1) {.smaller}

Install the main R packages with the statistical methods needed for this course:

```{r install1, eval=FALSE, echo=TRUE}
#Note: You might need to set .libPaths() to your local library path.
#Check your library paths:
.libPaths()

#To install packages using BiocManager with R (at least version "3.6") and enter:
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

#Use BiocManager to install many packages at the same time:
#Note: This will take a few minutes!
BiocManager::install(ask=FALSE, update=FALSE, c(
  #Unsupervised learning:
      "pheatmap","cluster","ConsensusClusterPlus","RColorBrewer",
  #High-throughput screening:
      "ISLR2","limma","edgeR","DESeq2", "matrixStats", "pasilla", 
  #Supervised learning:
      "glmnet",
  #Model validation:
      "c060", "M3C", "riskRegression", "styler", "htmltools"
))
```

## Let's get started (2)

Install additional R packages for data handling, data pre-processing and annotation, as well as some auxiliary packages that represent extensions to basic R:

```{r install2, eval=FALSE, echo=TRUE}
#Use BiocManager to install many packages at the same time:
BiocManager::install(ask=FALSE, update=FALSE, c(
  #Basic stats and bioinformatics packages:
      "Hmisc","psych","Biobase","genefilter","VennDiagram","boot",
      
  #Annotation and data pre-processing:
      "org.Hs.eg.db","hgu133plus2.db","AnnotationDbi",
      "biomaRt","GEOquery","Biostrings","gcrma","affy",
      "GenomicRanges","Gviz","rtracklayer","seqinr",
      "ISLR","airway", #Example data sets

  #Extensions to basic R (tidyverse, Rmarkdown etc):
      "fitdistrplus","tidyverse","reshape2",
      "data.table","optparse","rjson","yaml","dplR","readr",
      "logging","microbenchmark", "readxl"))
```

## Let's get started (3) {.smaller}

**Let's get started (3)** is a list of the R packages you will need for the various exercises in Simon's lectures. They are split into packages that can be installed with the standard `install.packages("packagename")` command, and those that must be installed with the `BiocManager::install("packagename")` command. See [here](https://intro2r.com/packages.html) for more information. Additionally, there are some other tasks listed afterwards that need to be taken care of.

## Required CRAN Packages (installed via `install.packages`)

```{r install3, eval=FALSE, echo=TRUE}
#For Code Development
if(!require(logging)){
  install.packages("logging")
}
if(!require(logger)){
  install.packages("logger")
}
if(!require(roxygen2)){
  install.packages("roxygen2")
}

if(!require(optparse)){
  install.packages("optparse")
}
if(!require(dplR)){
  install.packages("dplR")
}
if(!require(devtools)){
  install.packages("devtools") # for installing packages from GitHub
}
if(!require(tools)){
  install.packages("tools") # for package development
}
if(!require(BiocParallel)){
  install.packages("BiocParallel") # for package development
}


# For Biological Sequence Processing
if(!require(seqinr)){
  install.packages("seqinr")
}
if(!require(Biostrings)){
  install.packages("Biostrings")
}
if(!require(SummarizedExperiment)){
  install.packages("SummarizedExperiment")
}

# For accessing Cancer Datasets in the Gerstung paper
# if(!require(cgdsr)){
#   install.packages("cgdsr")
# }


# For reading/writing data
if(!require(data.table)){
  install.packages("data.table")
}

if(!require(rjson)){
install.packages("rjson")
}

if(!require(yaml)){
install.packages("yaml")
}


# For making pretty plots
if(!require(viridis)){
  install.packages("viridis")
}

if(!require(ggbump)){
  install.packages("ggbump")
}

if(!require(ggrepel)){
  install.packages("ggrepel")
}

#For PCA analysis & Stats
if(!require(ggfortify)){
install.packages("ggfortify")
}
if(!require(psych)){
install.packages("psych")
}
if(!require(Hmisc)){
  install.packages("Hmisc")
}
if(!require(Biostrings)){
  install.packages("e1071")
}  
if(!require(grid)){
  install.packages("grid")
} 
  
  
```

## Required BioConductor Packages (installed via `BiocManager::install`)

```{r BioConductor, echo=TRUE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(c(
  "rtracklayer","GenomicAlignments","pasillaBamSubset","Rsamtools",
  "airway","TxDb.Dmelanogaster.UCSC.dm3.ensGene","hgu133plus2.db","RBioinf"))
```

## Required GitHub Packages (installed via `install_github`)

```{r github, echo=TRUE}
#install_github("mg14/CoxHD/CoxHD")
```

## Let's get started (4) {.smaller}

Please also make sure that you can load the dataset that we will use throughout the course:

```{r loaddata, eval=FALSE, echo=TRUE}
#First, set the working directory to the directory where you have your
#data (available from Canvas) and where you want to save all your 
#analysis results. For example:
library(here)
# setwd("~/Dropbox/MF9155_teaching/2021")

#Then, load the RData objects:
load(here("input","Gerstung2015.RData"))
load(here("input","Gerstung2015_AML.Rdata"))

# # As an alternative you can create a path variable
# path <- "~/Dropbox/MF9155_teaching/2021/"
# #Then, load the RData objects:
# load(paste0(path,"Gerstung2015.RData"))
# load(paste0(path("Gerstung2015_AML.Rdata"))

#Check, that the loading of the data worked by confirming that you
#now have two new objects in your workspace called "gset" and "gsetAML":
ls()
```

## Let's get started (5) {.smaller}

### 5.1 Additional Software

During Simon's lectures it could be handy to have a text editor. You can use anything you want (Microsoft Word doesn't count as it doesn't create plain text files). Simon's suggestion: [Atom](https://atom.io/)

Also, access to a *Linux* command line window will be handy, but not necessary. If you have a MacOS then you are ready to go (just search for *term*) in *Spotlight Search*. If you are using Windows, you can get an Ubuntu app from the Windows store e.g., [this one](https://www.microsoft.com/en-us/p/ubuntu-2004-lts/9n6svws3rx71#activetab=pivot:overviewtab)
